Categories of Optics

### 2018-09-03 optics.tex
Put in a dot for the action everywhere

\begin{definition}
  Let $\C$ be a category and $(\M, \otimes, I)$ a monoidal category. An \emph{action of $\M$ on $\C$} is a monoidal functor $a : \M \to [\C, \C]$. For two objects $M \in \M$ and $A \in \C$, the action $a(M)(A)$ is abbreviated $M \act A$ or $MA$. The tensor product in $\M$ will also be written as juxtaposition, so $(NM)A \cong N(MA)$.% An action is \emph{lax} if $a$ is instead a lax monoidal functor.
\end{definition}

Given such an action, we define
\begin{align*}
  \Optic_\M((S, S'), (A, A')) := \int^{M \in \M} \C(S, MA) \times \C(MA', S')
\end{align*}
This subsumes the earlier definition, taking $\M = \C$ and having $\C$ act on itself via left-tensor:
\begin{align*}
a : \C &\to [\C, \C] \\
X &\mapsto X \otimes -
\end{align*}
We henceforth write this case as $\Optic_\otimes$, to emphasise the action on $\C$ that is used.

@@

\begin{definition}
Given two categories equipped with monoidal actions $(\M, \C)$ and $(\N, \D)$, a \emph{morphism of actions} is a monoidal functor $F^\bullet : \M \to N$ and a functor $F : \C \to \D$ that commutes with the actions, in the sense that there exist natural isomorphisms
  \begin{align*}
  \phi &: F(I A) \to (F^\bullet I)(FA) \\
  \phi_{M,A} &: F(MA) \to (F^\bullet M)(F A)
  \end{align*}
satisfying conditions analogous to those for a monoidal functor.

@@

\begin{remark}
We use $;$ to denote composition of $\C$ in diagrammatic order. The reason for this is that the coend relation can be applied simply by shifting the position of $\mid$ in a representative:
\begin{align*}
\rep{l;\phi A}{r} = \rep{l}{\phi A;r}
\end{align*}
%\todo{Should I switch the entire paper into using diagrammatic order?}
\end{remark}

Let $\Twoptic_\M(S, A)$ denote the set \[ \int^{M_1, M_2 \in \M} \C(S, M_1 A) \times \C(M_1 A, M_2 A) \times \C(M_2 A, S). \]
Using the universal property of the coend, we define three maps:
\begin{align*}
  \outside &: \Optic_\M(S, A) \to \C(S, S) \\
  \once, \twice &: \Optic_\M(S, A) \to \Twoptic_\M(S, A)
\end{align*}
by
\begin{align*}
  \outside(\rep{l}{r}) &= l;r \\
  \once(\rep{l}{r}) &= \repthree{l}{\id_{MA}}{r} \\
  \twice(\rep{l}{r}) &= \repthree{l}{r;l}{r}
\end{align*}

@@

\begin{proposition}\label{prop:lawful-lens-laws}
  A concrete lens given by $\fget$ and $\fput$ is lawful (in our sense) iff it obeys the three concrete lens laws.
\end{proposition}
\begin{proof}
  We begin by giving $\Twoptic_\times(S, A)$ the same treatment as $\Optic_\times(S, A)$. Using the universal property of the product and Yoneda reduction twice each, we have:
  \begin{align*}
  \Twoptic_\times(S, A)
  &= \int^{M_1, M_2 \in \C} \C(S, M_1 \times A) \times \C(M_1 \times A, M_2 \times A) \times \C(M_2 \times A, S) \\
  &\cong \int^{M_1, M_2 \in \C} \C(S, M_1) \times \C(S, A) \times \C(M_1 \times A, M_2 \times A) \times \C(M_2 \times A, S) \\
  &\cong \int^{M_2 \in \C} \C(S, A) \times \C(S \times A, M_2 \times A) \times \C(M_2 \times A, S) \\
  &\cong \int^{M_2 \in \C} \C(S, A) \times \C(S \times A, M_2) \times \C(S \times A, A) \times \C(M_2 \times A, S) \\
  &\cong \C(S, A) \times \C(S \times A, A) \times \C(S \times A \times A, S)
  \end{align*}
%  Call this last set $\conctwice_\times(S, A)$. Written equationally, the isomorphism $\Phi : \Twoptic_\times(S, A) \to \conctwice_\times(S, A)$ is given by:
  Written equationally, the isomorphism $\Phi : \Twoptic_\times(S, A) \to \C(S, A) \times \C(S \times A, A) \times \C(S \times A \times A, S)$ is given by:
  \begin{align*}
    \Phi(\repthree{l}{c}{r}) = (&\pi_2 l, \\
                                   &\pi_2 c (\pi_1 l \times A), \\
                                   &r(\pi_1 c (\pi_1 l \times A) \times A))
  \end{align*}
  Now suppose we are given a lens $p$ that corresponds concretely to $(\fget, \fput)$, so $p = \rep{[\id_S, \fget]}{\fput}$. Evaluating $\outside$ on this gives:
  \begin{align*}
    \outside(\rep{[\id_S, \fget]}{\fput}) = \fput [\id_S, \fget]
  \end{align*}
  so requiring $\outside(p) = \id_S$ is precisely the $\fget\fput$ law.

  We now have to slog through evaluating $\Phi(\once(p))$ and $\Phi(\twice(p))$.
  \begingroup
  \allowdisplaybreaks
  \begin{alignat*}{3}
    \Phi(\once(\rep{[\id_S, \fget]}{\fput})) &=
    \Phi(&&\repthree{[\id_S, \fget]}{\id_{S \times A}}{\fput}) \\
    &= (&&\pi_2 [\id_S, \fget], \\
    &&&\pi_2 \id_{S \times A} (\pi_1 [\id_S, \fget] \times A), \\
    &&& \fput (\pi_1 \id_{S \times A} (\pi_1 [\id_S, \fget] \times A) \times A) \quad) \\
    %
    &= (&&\fget, \\
    &&& \pi_2, \\
    &&& \fput \pi_{1,3} \quad) \\
    %%%%
    %%%%
    \Phi(\twice(\rep{[\id_S, \fget]}{\fput})) &=
    \Phi(&&\repthree{[\id_S, \fget]}{[\id_S, \fget]\fput}{\fput}) \\
    &= (&&\pi_2 [\id_S, \fget], \\
    &&&\pi_2 [\id_S, \fget]\fput (\pi_1 [\id_S, \fget] \times A), \\
    &&& \fput (\pi_1 [\id_S, \fget]\fput (\pi_1 [\id_S, \fget] \times A) \times A) \quad) \\
    %
    &= (&&\fget, \\
    &&& \fget \; \fput (\id_S \times A), \\
    &&& \fput (\fput (\id_S \times A) \times A) \quad) \\
    %
    &= (&&\fget, \\
    &&& \fget \; \fput, \\
    &&& \fput (\fput \times A) \quad)
  \end{alignat*}
  \endgroup
  So checking component-wise, $\Phi(\once(p))$ being equal to $\Phi(\twice(p))$ is exactly equivalent to the $\fput\fget$ and $\fput\fput$ laws holding.
\end{proof}

@@

\begin{proposition}
If $p = \rep{l}{r} : S \hto A$ is an optic such that $l$ and $r$ are mutual inverses, then $p$ is lawful.
\end{proposition}
\begin{proof}
The conditions are easy to check:
\begin{align*}
\outside(\rep{l}{r}) &= l;r = \id_S \\
\twice(\rep{l}{r})
&= \repthree{l}{r;l}{r} \\
&= \repthree{l}{\id_{MA}}{r} \\
&= \once(\rep{l}{r})
\end{align*}
\end{proof}

@@

\begin{corollary}\label{cor:tautological-lawful}
   For any two objects $A \in \C$ and $M \in \M$, the tautological optic $MA \hto A$ is lawful. \qed
\end{corollary}

\begin{proposition}
A costate $p : (S, S) \hto (I, I)$ corresponding to a morphism $f : S \to S$ via Proposition~\ref{prop:costates} is lawful iff $f = \id_S$.
\end{proposition}
\begin{proof}
The first law states that $\outside(p) = \id_S$, so if $p$ is lawful we have
\[  \id_S = \outside(\rep{\rho_S^{-1}}{f \rho_S}) =  f \rho_S\rho_S^{-1} = f \]
On the other hand, if $f  = \id_S$ then $\rep{\rho_S^{-1}}{\rho_S}$ is lawful because its components are mutual inverses.
\end{proof}

@@

  Suppose we have two lawful optics $\rep{l}{r} : R \hto S$ and $\rep{l'}{r'} : S \hto A$ with residuals $M$ and $N$ respectively. We must show that $\rep{(Ml')l}{r (Mr')}$ is also a lawful optic. Showing the first law is straightforward:
  \begin{align*}
    \outside(\rep{l; (Ml')}{(Mr') ; r})
    &= l ; Ml' ; Mr' ; r \\
    &= l ; M(l'r') ; r \\
    &= l ; M\id_{NA} ; r \\
    &= l ; r \\
    &= \id_S
  \end{align*}

  For the second law, we must show that
  \[ \repthree{ l;(Ml')}{(Mr'); r;l;(Ml')}{(Mr') ; r} = \repthree{l;(Ml')}{\id_{MNA}}{(Mr') ; r}. \]
  The idea is that, by the lawfulness of $\rep{l}{r}$ and $\rep{l'}{r'}$, there are chains of coend relations that prove
  \begin{align*}
  \repthree{l}{r;l}{r} &= \repthree{l}{\id_{MS}}{r} \\
  \repthree{l'}{r';l'}{r'} &= \repthree{l'}{\id_{NA}}{r'}
  \end{align*}
  The result is achieved by splicing these chains of relations together appropriately.

  So consider one of the generating relations $\repthree{l;\phi_S}{c}{r} = \repthree{l}{\phi_S ; c}{r}$ in $\Twoptic_\M(R, S)$. By the naturality of $\phi_A$, we calculate:
  \begin{align*}
    \repthree{l;\phi_S;(Ml')}{(Mr'); c ;(Ml')}{(Mr') ; r}
    &= \repthree{l;(Ml');\phi_{NA}}{(Mr'); c ;(Ml')}{(Mr') ; r} && \text{(functoriality)} \\
    &= \repthree{l;(Ml')}{\phi_{NA};(Mr'); c ;(Ml')}{(Mr') ; r} && \text{(coend relation)} \\
    &= \repthree{l;(Ml')}{(Mr');\phi_{S}; c ;(Ml')}{(Mr') ; r} && \text{(functoriality)}
  \end{align*}
  And similarly for the other generating relation, $\repthree{l}{c;\phi_S}{r} = \repthree{l}{c}{\phi_S; r}$.

  So indeed by replicating the same chain of relations that proves $\repthree{l}{r;l}{r} =\repthree{l}{\id_{MS}}{r}$, we see
  \begin{align*}
    \repthree{l;(Ml')}{(Mr');r;l;(Ml')}{(Mr') ; r} 
    &= \repthree{l;(Ml')}{(Mr');\id_{MS};(Ml')}{(Mr') ; r} \\
    &= \repthree{l;(Ml')}{M(r';l')}{(Mr') ; r}.
  \end{align*}

  Now that the $r;l$ in the center has been cleared away, we turn to the chain of relations proving $\repthree{l'}{r';l'}{r'} = \repthree{l'}{\id_{NA}}{r'}$. A generating relation $\repthree{l';\psi_A}{c'}{r'} = \repthree{l'}{\psi_A ; c'}{r'}$ in $\Twoptic_\M(S, A)$ implies that
  \begin{align*}
    \repthree{l;M(l';\psi_A)}{Mc'}{(Mr') ; r}
    &= \repthree{l;Ml';M\psi_A}{Mc'}{(Mr') ; r} \\
    &= \repthree{l;Ml'}{M\psi_A ; Mc'}{(Mr') ; r} \\
    &= \repthree{l;Ml'}{M(\psi_A;c')}{(Mr') ; r}
  \end{align*}
  And similarly for the generating relation on the other side. So again we can replicate the chain of relations proving $\repthree{l'}{r';l'}{r'} = \repthree{l'}{\id_{MA}}{r'}$ to show that
  \begin{align*}
    \repthree{l;(Ml')}{M(r';l')}{(Mr') ; r} 
    &= \repthree{l;(Ml')}{M(\id_{NA})}{(Mr') ; r} \\
    &= \repthree{l;(Ml')}{\id_{MNA}}{(Mr') ; r}
  \end{align*}
as required. We conclude that composition preserves lawfulness, so $\Lawful_\M$ is indeed a subcategory of $\Optic_\M$.
\end{proof}

@@

\begin{proposition}
  Suppose $F : (\M, \C) \to (\N, \D)$ is a morphism of actions. Then $\Optic(F) : \Optic_\M \to \Optic_\N$ restricts to a functor $\Lawful_\M \to \Lawful_\N$.
\end{proposition}
\begin{proof}
  If $p = \rep{l}{r}$ is lawful, then verifying the first equation is easy:
  \begin{align*}
  \outside(\Optic(F)(\rep{l}{r}))
  &= \outside\left(\rep{(Fl);\phi^{-1}_{M,A}}{\phi_{M,A'};(Fr)}\right) \\
  &= (Fl);\phi^{-1}_{M,A};\phi_{M,A'};(Fr)\\
  &= (Fl);(Fr)\\
  &= \id_{FS}
  \end{align*}
  where $\phi_{M,A'} : (F^\bullet M)(FA) \to F(MA)$ is the structure map that commutes $F$ with the actions.

  For the second equation, consider a generating relation $\repthree{l;\psi A}{c}{r} = \repthree{l}{\psi A ; c}{r}$. We can use the naturality of $\phi$ to show
  \begin{align*}
  \repthree{F(l;\psi A);\phi^{-1}_{M,A}}{\phi_{M,A};Fc;\phi^{-1}_{N,A}}{\phi_{N,A};Fr}
  &= \repthree{Fl;F(\psi A);\phi^{-1}_{M,A}}{\phi_{M,A};Fc;\phi^{-1}_{N,A}}{\phi_{N,A};Fr} \\
  &= \repthree{Fl;\phi^{-1}_{M',A};(F^\bullet \psi) A}{\phi_{M,A};Fc;\phi^{-1}_{N,A}}{\phi_{N,A};Fr} \\
  &= \repthree{Fl;\phi^{-1}_{M',A}}{(F^\bullet \psi) A;\phi_{M,A};Fc;\phi^{-1}_{N,A}}{\phi_{N,A};Fr} \\
  &= \repthree{Fl;\phi^{-1}_{M',A}}{\phi_{M',A};F(\psi A);Fc;\phi^{-1}_{N,A}}{\phi_{N,A};Fr} \\
  &= \repthree{Fl;\phi^{-1}_{M',A}}{\phi_{M',A};F(\psi A;c);\phi^{-1}_{N,A}}{\phi_{N,A};Fr}
  \end{align*}
  Similarly,
  \begin{align*}
    \repthree{Fl;\phi^{-1}_{M,A}}{\phi_{M,A};Fc;\phi^{-1}_{N,A}}{\phi_{N,A};F(\psi_A;r)}
    &= \repthree{Fl;\phi^{-1}_{M,A}}{\phi_{M,A};F(c;\psi A);\phi^{-1}_{N,A}}{\phi_{N,A};Fr}
  \end{align*}

  If $\rep{l}{r}$ is lawful, we can therefore replicate the chain of relations proving $\twice(\rep{l}{r}) = \once(\rep{l}{r})$ to show:
  \begin{align*}
  \twice(\Optic(F)(\rep{l}{r}))
  &= \repthree{Fl;\phi^{-1}_{M,A}}{\phi_{M,A};F(r;l);\phi^{-1}_{M,A}}{\phi_{M,A};Fr}  \\
  &= \repthree{Fl;\phi^{-1}_{M,A}}{\phi_{M,A};F(\id_{MA});\phi^{-1}_{M,A}}{\phi_{M,A};Fr}  \\
  &= \repthree{Fl;\phi^{-1}_{M,A}}{\id_{(F^\bullet M)(FA)}}{\phi_{M,A};Fr}  \\
  &= \once(\Optic(F)(\rep{l}{r}))
  \end{align*}
\end{proof}

@@

\begin{proposition}
  Let $\rep{l}{r} : S \hto A$ be an optic. If $l;r = \id_S$ and $r;l = \phi A$ for some $\phi : M \to M$ in $\M$, then $\rep{l}{r}$ is lawful.
\end{proposition}
\begin{proof}
  The statement $\outside(\rep{l}{r}) = l;r = \id_S$ is exactly the first law. And for the second, we verify:
  \begin{align*}
    \twice(\rep{l}{r})
    &= \repthree{l}{r;l}{r} \\
    &= \repthree{l}{\phi A}{r} && \text{($r;l = \phi A$)}\\
    &= \repthree{l ; (\phi A)}{\id_{MA}}{r} && \text{(coend relation)} \\
    &= \repthree{l;r;l}{\id_{MA}}{r} && \text{($r;l = \phi A$ again)}\\
    &= \repthree{l}{\id_{MA}}{r}  && \text{($l;r = \id_S$)}\\
    &= \once(\rep{l}{r})
  \end{align*}
\end{proof}

Even if $r;l = \phi A$ for some $\phi$, the same is not necessarily true for other representatives of the same optic. Let $\inside : \Optic_\M(S, A) \to \int^{M \in \M} \C(M A, M A)$ be the map induced by $\inside(\rep{l}{r}) = \langle r ; l \rangle$. We might ask that instead of requiring $r;l = \phi A$ exactly, we have $\langle r ; l \rangle = \langle \phi A \rangle$ in $\int^{M \in \M} \C(M A, M A)$. In fact, this is equivalent:

>>>>

\begin{definition}
  Let $\C$ be a category and $(\M, \otimes, I)$ a monoidal category. An \emph{action of $\M$ on $\C$} is a monoidal functor $a : \M \to [\C, \C]$. For two objects $M \in \M$ and $A \in \C$, the action $a(M)(A)$ is abbreviated $M \act A$. 
\end{definition}

Given such an action, we define
\begin{align*}
  \Optic_\M((S, S'), (A, A')) := \int^{M \in \M} \C(S, M \act A) \times \C(M \act A', S')
\end{align*}
This subsumes the earlier definition, taking $\M = \C$ and having $\C$ act on itself via left-tensor:
\begin{align*}
a : \C &\to [\C, \C] \\
X &\mapsto X \otimes -
\end{align*}
We henceforth write this case as $\Optic_\otimes$, to emphasise the action on $\C$ that is used.

@@

\begin{definition}
Given two categories equipped with monoidal actions $(\M, \C)$ and $(\N, \D)$, a \emph{morphism of actions} is a monoidal functor $F^\bullet : \M \to N$ and a functor $F : \C \to \D$ that commutes with the actions, in the sense that there exist natural isomorphisms
  \begin{align*}
  \phi &: F(I A) \to (F^\bullet I)(FA) \\
  \phi_{M,A} &: F(M \act A) \to (F^\bullet M) \act (F A)
  \end{align*}
satisfying conditions analogous to those for a monoidal functor.

@@

\begin{remark}
We use $;$ to denote composition of $\C$ in diagrammatic order. The reason for this is that the coend relation can be applied simply by shifting the position of $\mid$ in a representative:
\begin{align*}
\rep{l;(\phi \act A)}{r} = \rep{l}{(\phi \act A);r}
\end{align*}
%\todo{Should I switch the entire paper into using diagrammatic order?}
\end{remark}

Let $\Twoptic_\M(S, A)$ denote the set \[ \int^{M_1, M_2 \in \M} \C(S, M_1 \act A) \times \C(M_1 \act A, M_2 \act A) \times \C(M_2 \act A, S). \]
Using the universal property of the coend, we define three maps:
\begin{align*}
  \outside &: \Optic_\M(S, A) \to \C(S, S) \\
  \once, \twice &: \Optic_\M(S, A) \to \Twoptic_\M(S, A)
\end{align*}
by
\begin{align*}
  \outside(\rep{l}{r}) &= l;r \\
  \once(\rep{l}{r}) &= \repthree{l}{\id_{M\act A}}{r} \\
  \twice(\rep{l}{r}) &= \repthree{l}{r;l}{r}
\end{align*}

@@

\begin{proposition}\label{prop:lawful-lens-laws}
  A concrete lens given by $\fget$ and $\fput$ is lawful (in our sense) iff it obeys the three concrete lens laws.
\end{proposition}
\begin{proof}
  We begin by giving $\Twoptic_\times(S, A)$ the same treatment as $\Optic_\times(S, A)$. Using the universal property of the product and Yoneda reduction twice each, we have:
  \begin{align*}
  \Twoptic_\times(S, A)
  &= \int^{M_1, M_2 \in \C} \C(S, M_1 \times A) \times \C(M_1 \times A, M_2 \times A) \times \C(M_2 \times A, S) \\
  &\cong \int^{M_1, M_2 \in \C} \C(S, M_1) \times \C(S, A) \times \C(M_1 \times A, M_2 \times A) \times \C(M_2 \times A, S) \\
  &\cong \int^{M_2 \in \C} \C(S, A) \times \C(S \times A, M_2 \times A) \times \C(M_2 \times A, S) \\
  &\cong \int^{M_2 \in \C} \C(S, A) \times \C(S \times A, M_2) \times \C(S \times A, A) \times \C(M_2 \times A, S) \\
  &\cong \C(S, A) \times \C(S \times A, A) \times \C(S \times A \times A, S)
  \end{align*}
%  Call this last set $\conctwice_\times(S, A)$. Written equationally, the isomorphism $\Phi : \Twoptic_\times(S, A) \to \conctwice_\times(S, A)$ is given by:
  Written equationally, the isomorphism $\Phi : \Twoptic_\times(S, A) \to \C(S, A) \times \C(S \times A, A) \times \C(S \times A \times A, S)$ is given by:
  \begin{align*}
    \Phi(\repthree{l}{c}{r}) = (\quad&l;\pi_2, \\
                                       &(l;\pi_1 \times A);c;\pi_2, \\
                                       &((l;\pi_1 \times A);c;\pi_1 \times A);r \quad )
  \end{align*}
  Now suppose we are given a lens $p$ that corresponds concretely to $(\fget, \fput)$, so $p = \rep{[\id_S, \fget]}{\fput}$. Evaluating $\outside$ on this gives:
  \begin{align*}
    \outside(\rep{[\id_S, \fget]}{\fput}) = [\id_S, \fget];\fput 
  \end{align*}
  so requiring $\outside(p) = \id_S$ is precisely the $\fget\fput$ law.

  We now have to slog through evaluating $\Phi(\once(p))$ and $\Phi(\twice(p))$.
  \begingroup
  \allowdisplaybreaks
  \begin{alignat*}{3}
    \Phi(\once(\rep{[\id_S, \fget]}{\fput})) &=
    \Phi(&&\repthree{[\id_S, \fget]}{\id_{S \times A}}{\fput}) \\
    &= (&& [\id_S, \fget];\pi_2, \\
    &&& ( [\id_S, \fget];\pi_1 \times A);\id_{S \times A};\pi_2, \\
    &&& (( [\id_S, \fget];\pi_1 \times A);\id_{S \times A};\pi_1 \times A) ; \fput \quad) \\
    %
    &= (&&\fget, \\
    &&& \pi_2, \\
    &&& \pi_{1,3} ;\fput \quad) \\
    %%%%
    %%%%
    \Phi(\twice(\rep{[\id_S, \fget]}{\fput})) &=
    \Phi(&&\repthree{[\id_S, \fget]}{\fput;[\id_S, \fget]}{\fput}) \\
    &= (&& [\id_S, \fget];\pi_2, \\
    &&& ([\id_S, \fget];\pi_1  \times A);\fput;[\id_S, \fget];\pi_2, \\
    &&& (([\id_S, \fget];\pi_1  \times A);\fput;[\id_S, \fget];\pi_1 \times A);\fput \quad) \\
    %
    &= (&&\fget, \\
    &&& (\id_S \times A);\fput;\fget, \\
    &&& ((\id_S \times A);\fput  \times A);\fput \quad) \\
    %
    &= (&&\fget, \\
    &&& \fput;\fget, \\
    &&& (\fput \times A);\fput \quad)
  \end{alignat*}
  \endgroup
  So comparing component-wise, $\Phi(\once(p))$ being equal to $\Phi(\twice(p))$ is exactly equivalent to the $\fput\fget$ and $\fput\fput$ laws holding.
\end{proof}

@@

\begin{proposition}
If $p = \rep{l}{r} : S \hto A$ is an optic such that $l$ and $r$ are mutual inverses, then $p$ is lawful.
\end{proposition}
\begin{proof}
The conditions are easy to check:
\begin{align*}
\outside(\rep{l}{r}) &= l;r = \id_S \\
\twice(\rep{l}{r})
&= \repthree{l}{r;l}{r} \\
&= \repthree{l}{\id_{M\act A}}{r} \\
&= \once(\rep{l}{r})
\end{align*}
\end{proof}

@@

\begin{corollary}\label{cor:tautological-lawful}
   For any two objects $A \in \C$ and $M \in \M$, the tautological optic $M \act A \hto A$ is lawful. \qed
\end{corollary}

\begin{proposition}
A costate $p : (S, S) \hto (I, I)$ corresponding to a morphism $f : S \to S$ via Proposition~\ref{prop:costates} is lawful iff $f = \id_S$.
\end{proposition}
\begin{proof}
The first law states that $\outside(p) = \id_S$, so if $p$ is lawful we have
\[  \id_S = \outside(\rep{\rho_S^{-1}}{\rho_S;f}) =  \rho_S^{-1};\rho_S;f = f \]
On the other hand, if $f  = \id_S$ then $\rep{\rho_S^{-1}}{\rho_S}$ is lawful because its components are mutual inverses.
\end{proof}

@@

  Suppose we have two lawful optics $\rep{l}{r} : R \hto S$ and $\rep{l'}{r'} : S \hto A$ with residuals $M$ and $N$ respectively. We must show that $\rep{l;(M\act l')}{(M\act r');r}$ is also a lawful optic. Showing the first law is straightforward:
  \begin{align*}
    \outside(\rep{l; (M\act l')}{(M\act r') ; r})
    &= l ; (M \act l') ; (M \act r') ; r \\
    &= l ; (M \act l'r') ; r \\
    &= l ; (M \act \id_{NA}) ; r \\
    &= l ; r \\
    &= \id_S
  \end{align*}

  For the second law, we must show that
  \[ \repthree{ l;(M\act l')}{(M\act r'); r;l;(M\act l')}{(M\act r') ; r} = \repthree{l;(M\act l')}{\id_{MNA}}{(M\act r') ; r}. \]
  The idea is that, by the lawfulness of $\rep{l}{r}$ and $\rep{l'}{r'}$, there are chains of coend relations that prove
  \begin{align*}
  \repthree{l}{r;l}{r} &= \repthree{l}{\id_{MS}}{r} \\
  \repthree{l'}{r';l'}{r'} &= \repthree{l'}{\id_{NA}}{r'}
  \end{align*}
  The result is achieved by splicing these chains of relations together in the following way.

  Consider one of the generating relations $\repthree{l;(\phi \act S)}{c}{r} = \repthree{l}{(\phi \act S) ; c}{r}$ in $\Twoptic_\M(R, S)$. By the naturality of $\phi_A$, we calculate:
  \begin{align*}
    \repthree{l;(\phi \act S);(M \act l')}{(M\act r'); c ;(M\act l')}{(M\act r') ; r}
    &= \repthree{l;(M \act l');(\phi \act NA)}{(M \act r'); c ;(M \act l')}{(M \act r') ; r} && \text{(functoriality)} \\
    &= \repthree{l;(M \act l')}{\phi_{NA};(M \act r'); c ;(M \act l')}{(M \act r') ; r} && \text{(coend relation)} \\
    &= \repthree{l;(M \act l')}{(M \act r');\phi_{S}; c ;(M \act l')}{(M \act r') ; r} && \text{(functoriality)}
  \end{align*}
  And similarly for the other generating relation, $\repthree{l}{c;\phi_S}{r} = \repthree{l}{c}{\phi_S; r}$.

  So indeed by replicating the same chain of relations that proves $\repthree{l}{r;l}{r} =\repthree{l}{\id_{M \act S}}{r}$, we see
  \begin{align*}
    \repthree{l;(M \act l')}{(M \act r');r;l;(M \act l')}{(M \act r') ; r} 
    &= \repthree{l;(M \act l')}{(M \act r');\id_{M \act S};(M \act l')}{(M \act r') ; r} \\
    &= \repthree{l;(M \act l')}{(M \act r';l')}{(M \act r') ; r}.
  \end{align*}

  Now that the $r;l$ in the center has been cleared away, we turn to the chain of relations proving $\repthree{l'}{r';l'}{r'} = \repthree{l'}{\id_{NA}}{r'}$. A generating relation $\repthree{l';\psi_A}{c'}{r'} = \repthree{l'}{\psi_A ; c'}{r'}$ in $\Twoptic_\M(S, A)$ implies that
  \begin{align*}
    \repthree{l;(M \act l';\psi_A)}{M \act c'}{(M \act r') ; r}
    &= \repthree{l;(M\act l');(M \act \psi \act A}{M \act c'}{(M \act r') ; r} \\
    &= \repthree{l;(M \act l')}{(M \act \psi \act A) ; (M \act c')}{(M\act r') ; r} \\
    &= \repthree{l;(M \act l')}{M \act ((\psi \act A);c')}{(M \act r') ; r}
  \end{align*}
  And similarly for the generating relation on the other side. So again we can replicate the chain of relations proving $\repthree{l'}{r';l'}{r'} = \repthree{l'}{\id_{M\act A}}{r'}$ to show that
  \begin{align*}
    \repthree{l;(M \act l')}{(M \act r';l')}{(M \act r') ; r} 
    &= \repthree{l;(M \act l')}{M(\id_{N\act A})}{(M \act r') ; r} \\
    &= \repthree{l;(M \act l')}{\id_{M \act N \act A}}{(M \act r') ; r}
  \end{align*}
as required. We conclude that composition preserves lawfulness, so $\Lawful_\M$ is indeed a subcategory of $\Optic_\M$.
\end{proof}

@@

\begin{proposition}
  Suppose $F : (\M, \C) \to (\N, \D)$ is a morphism of actions. Then $\Optic(F) : \Optic_\M \to \Optic_\N$ restricts to a functor $\Lawful_\M \to \Lawful_\N$.
\end{proposition}
\begin{proof}
  If $p = \rep{l}{r}$ is lawful, then verifying the first equation is easy:
  \begin{align*}
  \outside(\Optic(F)(\rep{l}{r}))
  &= \outside\left(\rep{(Fl);\phi^{-1}_{M,A}}{\phi_{M,A'};(Fr)}\right) \\
  &= (Fl);\phi^{-1}_{M,A};\phi_{M,A'};(Fr)\\
  &= (Fl);(Fr)\\
  &= \id_{FS}
  \end{align*}
  where $\phi_{M,A'} : (F^\bullet M) \act (FA) \to F(M \act A)$ is the structure map that commutes $F$ with the actions.

  For the second equation, consider a generating relation $\repthree{l;(\psi \act A)}{c}{r} = \repthree{l}{(\psi \act A) ; c}{r}$. We can use the naturality of $\phi$ to show
  \begin{align*}
  \repthree{F(l;\psi \act A);\phi^{-1}_{M,A}}{\phi_{M,A};Fc;\phi^{-1}_{N,A}}{\phi_{N,A};Fr}
  &= \repthree{Fl;F(\psi \act A);\phi^{-1}_{M,A}}{\phi_{M,A};Fc;\phi^{-1}_{N,A}}{\phi_{N,A};Fr} \\
  &= \repthree{Fl;\phi^{-1}_{M',A};(F^\bullet \psi) \act A}{\phi_{M,A};Fc;\phi^{-1}_{N,A}}{\phi_{N,A};Fr} \\
  &= \repthree{Fl;\phi^{-1}_{M',A}}{(F^\bullet \psi) \act A;\phi_{M,A};Fc;\phi^{-1}_{N,A}}{\phi_{N,A};Fr} \\
  &= \repthree{Fl;\phi^{-1}_{M',A}}{\phi_{M',A};F(\psi \act A);Fc;\phi^{-1}_{N,A}}{\phi_{N,A};Fr} \\
  &= \repthree{Fl;\phi^{-1}_{M',A}}{\phi_{M',A};F(\psi \act A;c);\phi^{-1}_{N,A}}{\phi_{N,A};Fr}
  \end{align*}
  Similarly,
  \begin{align*}
    \repthree{Fl;\phi^{-1}_{M,A}}{\phi_{M,A};Fc;\phi^{-1}_{N,A}}{\phi_{N,A};F((\psi \act A);r)}
    &= \repthree{Fl;\phi^{-1}_{M,A}}{\phi_{M,A};F(c;(\psi \act A));\phi^{-1}_{N,A}}{\phi_{N,A};Fr}
  \end{align*}

  If $\rep{l}{r}$ is lawful, we can therefore replicate the chain of relations proving $\twice(\rep{l}{r}) = \once(\rep{l}{r})$ to show:
  \begin{align*}
  \twice(\Optic(F)(\rep{l}{r}))
  &= \repthree{Fl;\phi^{-1}_{M,A}}{\phi_{M,A};F(r;l);\phi^{-1}_{M,A}}{\phi_{M,A};Fr}  \\
  &= \repthree{Fl;\phi^{-1}_{M,A}}{\phi_{M,A};F(\id_{M \act A});\phi^{-1}_{M,A}}{\phi_{M,A};Fr}  \\
  &= \repthree{Fl;\phi^{-1}_{M,A}}{\id_{(F^\bullet M) \act (FA)}}{\phi_{M,A};Fr}  \\
  &= \once(\Optic(F)(\rep{l}{r}))
  \end{align*}
\end{proof}

@@

\begin{proposition}
  Let $\rep{l}{r} : S \hto A$ be an optic. If $l;r = \id_S$ and $r;l = \phi \act A$ for some $\phi : M \to M$ in $\M$, then $\rep{l}{r}$ is lawful.
\end{proposition}
\begin{proof}
  The statement $\outside(\rep{l}{r}) = l;r = \id_S$ is exactly the first law. And for the second, we verify:
  \begin{align*}
    \twice(\rep{l}{r})
    &= \repthree{l}{r;l}{r} \\
    &= \repthree{l}{\phi \act A}{r} && \text{($r;l = \phi \act A$)}\\
    &= \repthree{l ; (\phi \act A)}{\id_{M\act A}}{r} && \text{(coend relation)} \\
    &= \repthree{l;r;l}{\id_{M\act A}}{r} && \text{($r;l = \phi \act A$ again)}\\
    &= \repthree{l}{\id_{M\act A}}{r}  && \text{($l;r = \id_S$)}\\
    &= \once(\rep{l}{r})
  \end{align*}
\end{proof}

Even if $r;l = \phi A$ for some $\phi$, the same is not necessarily true for other representatives of the same optic. Let $\inside : \Optic_\M(S, A) \to \int^{M \in \M} \C(M \act  A, M \act  A)$ be the map induced by $\inside(\rep{l}{r}) = \langle r ; l \rangle$. We might ask that instead of requiring $r;l = \phi A$ exactly, we have $\langle r ; l \rangle = \langle \phi A \rangle$ in $\int^{M \in \M} \C(M \act  A, M \act  A)$. In fact, this is equivalent:

\begin{proposition}\label{prop:onthenose}
  Suppose $p : S \hto A$ satisfies $\outside(p) = \id_S$ and $\inside(p) = \langle \phi A \rangle$. Then there exists a representative $\rep{l}{r}$ such that $r;l = \psi A$ on the nose for some (possibly different) $\psi : M \to M$.
\end{proposition}
\begin{proof}
  The generating relation for $\int^{M \in \M} \C(M \act  A, M \act  A)$ is
  \[ \langle f; \phi_A \rangle = \langle \phi_A; f \rangle \]
  whenever $f : N \act A \to M \act A$ and $\phi : M \to N$. This relation $f; (\phi \act A) \rightsquigarrow (\phi \act A); f$ is not likely to be symmetric or transitive in general. Note that if $f; (\phi \act A) \rightsquigarrow (\phi \act A); f$ then $f ;(\phi \act A) ;f; (\phi \act A) \rightsquigarrow (\phi \act A) ;f ;(\phi \act A); f$. More generally, if $f \rightsquigarrow g$ then $f^n \rightsquigarrow g^n$ for any $n$.
